National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2023-11-21 analysis/paper/paper.Rmd
ongoing work on paper

Collared females are also used in most estimates of recruitment. Recruitment surveys involve locating collared individuals by aircraft and then surveying the entire group of caribou to estimate the ratio of calves to cows in the group [@nbckc_nbckc_2022]. Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below). Note that occasionally young bulls can be misidentified as cows (parameters $\dot{u}$ and $u$ below; refs? ), and there is also a possibility that some calves may be missed during aerial composition surveys (parameters $\dot{z}$ and $z$ below). We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details). Telemetry and aerial survey data from these methods can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for monitoring population change of many caribou herds [e.g. @decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019].

@@

To make this work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios. Where appropropriate, we borrowed and built on code from an app developed by Eacker et al. [-@eacker_web-based_2019]. Our app allows
users to provide existing local data or simulate data based on national
models as shown above, define an anthropogenic disturbance scenario and
make projections. The results shown are similar to Figure
\@ref(fig:examplePredictions) and allow for comparisons to the national
model. The app allows advanced users to adjust most model parameters (supplemental Table A1). Default parameter values are set to allow novice users to get reasonable results quickly. Most of the code needed to run the app is in an R package, caribouMetrics, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of contexts (e.g. Frances SpaDES paper, Dyson preprint, syncrosim reference). The caribouMetrics package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>. Find documentation and vignettes on the package website
<https://landscitech.github.io/caribouMetrics/>. The shiny app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>, and code required to run analyses and produce figures in the paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.

>>>>

Collared females are also used in most estimates of recruitment. Recruitment surveys involve locating collared individuals by aircraft and then surveying the entire group of caribou to estimate the ratio of calves to cows in the group [@nbckc_nbckc_2022]. Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below). However, a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020]. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below; refs? ), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020]. We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details). Telemetry and aerial survey data from these methods can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for monitoring population change of many caribou herds [e.g. @decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019].

@@

To make this work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios. Where appropropriate, we borrowed and built on code from an app developed by Eacker et al. [-@eacker_web-based_2019]. Our app allows
users to provide existing local data or simulate data based on national
models as shown above, define an anthropogenic disturbance scenario and
make projections. The results shown are similar to Figure
\@ref(fig:examplePredictions) and allow for comparisons to the national
model. The app allows advanced users to adjust many model parameters (supplemental Table A1). Default parameter values are set to allow novice users to get reasonable results quickly. Most of the code needed to run the app is in an R package, caribouMetrics, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g. @dyson_existing_2022; @stewart_climate-informed_2023]. The caribouMetrics package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>. Find documentation and vignettes on the package website
<https://landscitech.github.io/caribouMetrics/>. The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>, and code required to run analyses and produce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.

### 2023-11-21 analysis/paper/supplement.Rmd
ongoing work on paper

```{r PriorKSDists, fig.cap="Similarity of the prior survival distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022]. The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\\sigma^2_S$ in Table 2). Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
knitr::include_graphics(here::here("figs/PriorKSDists.png"))

@@

```{r PriorAllFits, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}

@@

```{r PriorLowFits, fig.cap="Posterior predictive survival distributions from the Bayesian model ('local') assuming intensive monitoring (15 years of 60 collars per year) of populations from the 2.5 \\% quantile of the national model. In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and  anthropogenic disturbance is 0. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates. When the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is too low (<5) the insufficiently vague priors substantially influence the posterior distributions, yielding poor local parameter estimates.", cache=TRUE}

knitr::include_graphics(here::here("figs/SurvsurvInterceptLowsSigmaSD0.03.png"))
```

@@

Prior parameter values should be set so that, given no monitoring data, the prior survival distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance. This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)). Priors should also be vague enough so that, given intensive monitoring of populations where survival is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 60 collars per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local survival, rather than priors from the national model. Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (SigmaMean=0.034784, SigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).

```{r PriorKSDistsRec, fig.cap="Similarity of the prior recruitment distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022]. The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (rIntSEMod: $\\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\\sigma^2_R$ in Table 2). Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
knitr::include_graphics(here::here("figs/PriorKSDistsRec.png"))
```

```{r PriorAllFitsRec, fig.cap="Differences between prior predictive recruitment distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, the standard deviation of the random effect prior (SigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
knitr::include_graphics(here::here("figs/RecInterceptAllrSigmaSD0.22.png"))
```

```{r PriorHighFitsRec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model ('local') assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5% quantile of the national model.  In these examples, the standard deviation the random effect prior (SigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and  anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals, black dots show true recruitment rates from the simulated population, and open triangles show observed recruitment rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecInterceptHighrSigmaSD0.22.png"))
```

@@

```{r PriorHighFitsAnthro90Rec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model, assuming intensive monitoring (15 years of 60 collars per year) of populations from the 97.5 \\% quantile of the national model. In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (rSlopeMod=0) through expected (rSlopeMod=1) to double the expected slope (rSlopeMod=2). All considered multipliers on the prior standard deviation of the recruitment slope >1 (rSlopeSEMod: $\\beta^R_a$ in Table 2) are sufficiently vague to reflect local variation among populations. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecSlopeHigh.png"))
```

Similarly, prior parameter values should be set so that, given no monitoring data, the prior recruitment distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance. This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (SigmaMean & SigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)). Priors should also be vague enough so that, given intensive monitoring of populations where recruitment is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 9*60 cows per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local recruitment, rather than priors from the national model. Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (SigmaMean=0.23, SigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).

>>>>

```{r PriorKSDists, fig.cap="Similarity of the prior survival distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022]. The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\\sigma^2_S$ in Table 2). Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
knitr::include_graphics(here::here("figs/PriorKSDists.png"))

@@

```{r PriorAllFits, fig.cap="Differences between prior predictive survival distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, the standard deviation of the random effect prior (sSigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}

@@

```{r PriorLowFits, fig.cap="Posterior predictive survival distributions from the Bayesian model ('local') assuming intensive monitoring (15 years of 60 collars per year) of populations from the 2.5 \\% quantile of the national model. In these examples, the standard deviation of the random effect prior (sSigmaSD: $\\sigma^2_S$ in Table 2) is 0.03, and  anthropogenic disturbance is 0. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates. When the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\\beta^S_0$ in Table 2) is too low (<5) the insufficiently vague priors substantially influence the posterior distributions, yielding poor local parameter estimates.", cache=TRUE}

knitr::include_graphics(here::here("figs/SurvInterceptLowsSigmaSD0.03.png"))
```

@@

Prior parameter values should be set so that, given no monitoring data, the prior survival distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance. This condition is met when the multiplier on the prior standard deviation of the survival intercept (sIntSEMod: $\beta^S_0$ in Table 2) is 5 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (sSigmaMean & sSigmaSD, $\sigma^2_S$ in Table 2) (Figs \@ref(fig:PriorKSDists) and \@ref(fig:PriorAllFits)). Priors should also be vague enough so that, given intensive monitoring of populations where survival is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 60 collars per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local survival, rather than priors from the national model. Setting the mean and standard deviations of the prior distribution of the random effect of year on survival to moderately low values (sSigmaMean=0.034784, sSigmaSD=0.03, $\sigma^2_S$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDists), \@ref(fig:PriorAllFits) and \@ref(fig:PriorLowFits)).

```{r PriorKSDistsRec, fig.cap="Similarity of the prior recruitment distribution to the distribution of simulated outcomes from the national model measured by Kolmogorov-Smirnov distance [ks.test function in @r_core_team_r_2022]. The distributions are similar when the multiplier on the prior standard deviation of the survival intercept (rIntSEMod: $\\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\\sigma^2_R$ in Table 2). Anthropogenic disturbance is set to 0 in these scenarios." , cache=TRUE}
knitr::include_graphics(here::here("figs/PriorKSDistsRec.png"))
```

```{r PriorAllFitsRec, fig.cap="Differences between prior predictive recruitment distributions from the Bayesian model ('local') and simulated outcomes from the national model ('national'), assuming no local monitoring. In these examples, the standard deviation of the random effect prior (rSigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals ('local'), and the 2.5th and 97.5th percentiles for 1000 sample populations from the national model.",  cache=TRUE}
knitr::include_graphics(here::here("figs/RecInterceptAllrSigmaSD0.22.png"))
```

```{r PriorHighFitsRec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model ('local') assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5% quantile of the national model.  In these examples, the standard deviation the random effect prior (rSigmaSD: $\\sigma^2_R$ in Table 2) is 0.22, and  anthropogenic disturbance is 0. Bands show 95% posterior predictive intervals, black dots show true recruitment rates from the simulated population, and open triangles show observed recruitment rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecInterceptHighrSigmaSD0.22.png"))
```

@@

```{r PriorHighFitsAnthro90Rec, fig.cap="Posterior predictive recruitment distributions from the Bayesian model, assuming intensive monitoring (9*60 cows per year for 15 years) of populations from the 97.5 \\% quantile of the national model. In these examples, anthropogenic disturbance is 90\\%, and the effect of anthropogenic disturbance varies from none (rSlopeMod=0) through expected (rSlopeMod=1) to double the expected slope (rSlopeMod=2). All considered multipliers on the prior standard deviation of the recruitment slope >1 (rSlopeSEMod: $\\beta^R_a$ in Table 2) are sufficiently vague to reflect local variation among populations. Bands show 95 \\% posterior predictive intervals, black dots show true survival rates from the simulated population, and open triangles show observed survival rates.", cache=TRUE}
knitr::include_graphics(here::here("figs/RecSlopeHigh.png"))
```

Similarly, prior parameter values should be set so that, given no monitoring data, the prior recruitment distribution is similar to the distribution of simulated outcomes from the national model in the case of no anthropogenic disturbance. This condition is met when the multiplier on the prior standard deviation of the recruitment intercept (rIntSEMod: $\beta^R_0$ in Table 2) is 4 and the mean and standard deviations of the prior distribution of the random effect of year on survival are low  (rSigmaMean & rSigmaSD, $\sigma^2_R$ in Table 2) (Figs \@ref(fig:PriorKSDistsRec) and \@ref(fig:PriorAllFitsRec)). Priors should also be vague enough so that, given intensive monitoring of populations where recruitment is higher or lower than the national average ('High' and 'Low' scenarios: 15 years of 9*60 cows per year from 97.5% and 2.5% quantiles of the national model), posteriors reflect local recruitment, rather than priors from the national model. Setting the mean and standard deviations of the prior distribution of the random effect of year on recruitment to moderately low values (rSigmaMean=0.23, rSigmaSD=0.22, $\sigma^2_R$ in Table 2) allows both conditions to be met (Figs \@ref(fig:PriorKSDistsRec), \@ref(fig:PriorAllFitsRec) and \@ref(fig:PriorHighFitsRec)).
